---
title: 'Lecture 6: Linear Regression'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.align = "center")
library(tidyverse)
```

In [Lecture 4](lecture4.html), we looked at predicting an MLB player's 2015 batting average using their 2014 batting average. We did so by binning the data and looking at the average within each bin. What if we wanted to make the bins infinitesimally small? This brings us to the ideas of *correlation* and the *regression method*.

We again load the data (if you don't have it, you can download it [here](data/batting_2014_2015.RData)) and use the `pivot_wider` function to separate the 2014 and 2015 batting averages into separate columns. 

```{r load batting_2014_2015, message = FALSE}
load("data/batting_2014_2015.RData")
batting_2014_2015 <- batting_2014_2015 %>%
  pivot_wider(
    names_from = yearID,
    values_from = BA,
    names_prefix = "BA_"
  )
head(batting_2014_2015)
```

Again we plot the 2014 batting averages against the 2015 batting averages:

```{r plot batting_2014_2015}
g <- ggplot(data = batting_2014_2015) +
  geom_point(aes(x = BA_2014, y = BA_2015)) + 
  labs(x = "2014 Batting Average", y = "2015 Batting Average")
g
```


### Correlation

We can see a positive trend between the 2014 and 2015 batting averages. How can we quantify the relationship between the 2014 and 2015 data? 

The **correlation coefficient**, *r*, is one way to summarize the dependence between two seasons with one number. *r* is a standardized measure of the *linear* dependence between two variables (usually called $x$ and $y$) and can take values between -1 and +1.  

If $r = 1$, then the points all lie on a line with positive slope. If $r = -1$, the points all lie on a line with negative slope. 

Loosely speaking, we can think of correlation as a measure of how well a line fits our data. [Let's get some practice!](http://guessthecorrelation.com/)

We can calculate the correlation between two variables in `R` using the function `cor()`. Note that `cor()` is a Base R function and does not work with the other tidyverse functions, so we cannot pipe our data into it. Instead, we need to specifically point to each variable within the dataset, using the dollar sign symbol we've seen a couple times before: `our_dataset$our_variable`. This is equivalent to the tidyverse notation for doing `our_dataset %>% select(our_variable)`.

The correlation between the 2014 and 2015 batting averages is:

```{r cor}
cor(batting_2014_2015$BA_2014, batting_2014_2015$BA_2015)
```

We have a moderate amount of correlation between 2014 and 2015 batting averages, but there is still a lot of noise. 

It is important to remember that correlation is only a measure of *linear* dependence between two variables. The below example shows data where $y = x^2$ exactly. 

```{r quadratic plot, echo = F}
set.seed(1234)
x <- runif(50, -1, 1)
y <- x^2
dat <- data.frame(x = x, y = y)
eg <- ggplot(data = dat, aes(x, y)) + geom_point() + labs(title = expression(paste("y = ", x^2)))
eg
```

Even though $x$ and $y$ are dependent, $r = -0.1$, indicating a weak linear dependence. If we were to draw a line to fit this data, it would be quite flat!

```{r quadratic plot line, echo = F}
fit <- summary(lm(y~x, data = dat))
a  <- fit$coefficients[1,1]
b <- fit$coefficients[2,1]
eg <- eg + geom_abline(intercept = a, slope = b, color = "red")
eg
```

Now let's go back to the 2014 and 2015 batting data. To visualize our correlation, we draw the *line of best fit* through our data.  In this example, the line of best fit has $y$-intercept = 0.141 and slope = 0.485 (we'll talk about how we find this values later). 

```{r slope intercept, echo=FALSE}
fit <- summary(lm(BA_2015~BA_2014, data = batting_2014_2015))
a_best   <- fit$coefficients[1,1]
b_best   <- fit$coefficients[2,1]
```

```{r slope intercept plot}
g <- g + geom_abline(intercept = 0.141, slope = 0.485, color = "red") + 
  labs(title = "y = 0.141 + 0.485x")
g
```

**Question:** Which feature of the best fit line, $a$ or $b$, quantifies the linear dependence between 2014 and 2015 FG%?

How does correlation relate to our line of best fit? This brings us to the *regression method*.


### The Regression Method

If we standardize both datasets, $x$ and $y$, the correlation is the *slope* of the line of best fit:

$$\frac{y - \bar{y}}{sd(y)} = r \times \frac{x - \bar{x}}{sd(x)}$$

We can unpack this equation to get the formula for our unstandardized line of best fit:

$$y = a + bx$$
where $a = \bar{y} - b\bar{x}$ and $b = r \times sd(y)/sd(x)$.

Now that we have our regression line, we can predict a future $y$ value given we know $x$. For example, if we know that a player's 2014 batting average was 0.31, we predict that their 2015 batting average will be:
$$\widehat{y} = 0.141 + 0.485 \times 0.31 = 0.291.$$
Note that we use $\widehat{y}$ for a predicted value of $y$; we can think of this as $\widehat{y} = E[Y|x]$.

```{r plot predicted y, echo = F, warnings = FALSE}
g <- ggplot(data = batting_2014_2015) + 
  geom_point(aes(x = BA_2014, y = BA_2015)) + 
  labs(x = "2014 Batting Average", y = "2015 Batting Average") + 
  geom_abline(intercept = 0.141, slope = 0.485, color = "red") +
  labs(title = "y = 0.141 + 0.485x") +
  geom_segment(aes(x = 0.19, y = 0.291, yend = 0.291, xend = 0.31), linetype = "dashed", color = "dodgerblue") +
  geom_segment(aes(x = 0.31, y = 0.20, yend = 0.291, xend = 0.31), linetype = "dashed", color = "dodgerblue") +
  geom_point(aes(x = 0.31, y = 0.291), color = "dodgerblue", size = 2) +
  ylim(0.2, 0.35) +
  xlim(0.19, 0.355) +
  coord_cartesian(expand = FALSE)
g
```

**Question:** What is the interpretation of the slope coefficient, $b$?


### Fitting linear models

To find the regression coefficients, $a$ and $b$, we use the function `lm()`. The usage of `lm()` is as follows:

```{r lm fn}
model <- lm(BA_2015 ~ BA_2014, data = batting_2014_2015)
```

The first argument in the `lm()` call is called a `formula`. This takes input `y ~ x`, where `y` is our response and `x` is our predictor or covariate.  In the `lm()` call above, the column `BA_2015` is our response and `BA_2014` is our predictor. 

The second argument in `lm()` is where we specifiy our data: `batting_2014_2015`. `R` then looks for the columns `BA_2014` and `BA_2015` in the given dataset to calculate the regression coefficients.

Our new object, `fit`, contains a lot of information about our regression line. At the moment, we just want the coefficients. We can access the coefficients as follows:
```{r get coefficients}
model$coefficients
```

We can plot the data by feeding these coefficients to `geom_abline()`, however this approach only works for linear regression. A more efficient way is to use `predict()` to add predicted values from the model to our dataset, then add these points to our plot with `geom_line()`. We can do this as below:
```{r plot predictions, warning = FALSE }
batting_2014_2015 <- batting_2014_2015 %>%
  mutate(pred = predict(model))

ggplot(data = batting_2014_2015) +
  geom_point(aes(x = BA_2014, y = BA_2015)) + 
  geom_line(aes(x = BA_2014, y = pred), color = "lightcoral", size = 1) +
  labs(x = "2014 Batting Average", y = "2015 Batting Average") +
  labs(title = "Batting Average w/ Regression Line")


```


### The `modelr` package

`modelr` is another package that is part of the `tidyverse`. It has a number of useful functions for linear regression models.

```{r load modelr}
#install.packages("modelr")
library(modelr)
```

We can use the function `rsquare` to get the square of the correlation, $r^2$. The quantity $r^2$ is the *proportion of variance* explained by the linear model. The first argument of the `rsquare` function is the output `fit` from our linear model function `lm`. The second argument is our original dataset, `batting_2014_2015`:

```{r rsquare}
rsquare(fit, batting_2014_2015)
```


#### Adding Predictions and Residuals

We can also use `modelr` to add predictions and residuals to our original dataset:

```{r add predictions residuals}
batting_2014_2015 <- batting_2014_2015 %>% 
  add_predictions(model = fit, type = "response", var = "pred") %>%
  add_residuals(model = fit, var = "resid")
batting_2014_2015
```

In `add_predictions`, we have to specify three arguments: (i) the model, which is just the output of `lm()`, (ii) the type, which will always be "response" for this course, and (iii) the name of the column we want our predictions to be stored in.
We similarly have to specify a `model` and `var` argument in `add_residual`. 

Using these new columns, we can create the original plot and add the residual lengths:

```{r plot residuals}
g <- ggplot(batting_2014_2015) +
  geom_segment(aes(x = BA_2014, xend = BA_2014, y = BA_2015, yend = pred), color = "dodgerblue") + 
  geom_point(aes(x = BA_2014, y = BA_2015)) + 
  geom_abline(intercept = 0.1409779, slope = 0.4851417, color = "red")
g
```


### Predict

So far we have looked at the predicted values from our *training* data; that is, the data we used to fit our linear model. Suppose we have more players' batting averages from 2014 but we do not have their batting averages from 2015. We can once again use the `predict` function to obtain values for these new players based on our model. 

In the below code, we enter the players' 2014 batting averages as the tibble `new_data`. We then use the function `predict`. The first argument of `predict` is the fitted model, `model`. The second argument is `new_data`.

```{r new_data predict}
new_data <- tibble(BA_2014 = c(0.241, 0.31, 0.265))
new_pred <- predict(model, new_data)
new_pred
```

We can also add these predictions to `new_data` using `add_predictions`:

```{r new_data add_predictions}
new_data <- new_data %>%
  add_predictions(model)
new_data
```

We add these predictions to our plot:

```{r plot new predictions}
ggplot(batting_2014_2015) +
  geom_point(aes(x = BA_2014, y = BA_2015)) +
  geom_line(aes(x = BA_2014, y = pred), color = "lightcoral", size = 1) +
  geom_point(data = new_data, mapping = aes(x = BA_2014, y = pred), color = "dodgerblue")
```

For more details on modelling using the `tidyverse`, you can look at [Chapter 23](http://r4ds.had.co.nz/model-basics.html) of the book *R for Data Science* by Hadley Wickham and Garrett Grolemund. It's also a great reference in general!
