---
title: "Super Bowl Analysis with nflscrapR"
output: html_document
---

```{r nfl-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "", prompt = TRUE, collapse = TRUE, tidy=FALSE, fig.align = "center")
library(tidyverse)
```

## Getting the data

In this example analysis, we're going to examine Super Bowl LII to gain insight into
how the Philadelphia Eagles beat the New England Patriots (last year's Super Bowl 
was pretty boring...)

The first step is to load the data. Although we can use the [`nflscrapR` package](https://github.com/maksimhorowitz/nflscrapR)
to do this, we're going to save time and access the repository of files that
are already available to load and analyze from [here](https://github.com/ryurko/nflscrapR-data).
We're going to read in the entire set of play-by-play data from the post-season for the 2017-2018 NFL season.

```{r load-pbp-data}
post_pbp_2017 <- read_csv("https://raw.githubusercontent.com/ryurko/nflscrapR-data/master/play_by_play_data/post_season/post_pbp_2017.csv")
```

This has loaded the entire play-by-play from the 2017 post-season (including the Pro Bowl!). 
We're going to filter this data down to the Super Bowl only, which has the `game_id` of 2018020400
And to make this simpler, we're going to `select` only a subset of the columns to work with

```{r super-bowl-data}
super_bowl_pbp <- post_pbp_2017 %>%
  # Only grab the Super Bowl
  filter(game_id == 2018020400) %>%
  # First grab context about the play:
  select(posteam, defteam, drive, qtr, down, ydstogo, yardline_100,
         half_seconds_remaining, score_differential,
         # Next data on the result:
         desc, play_type, yards_gained, sp,
         # Finally the advanced metrics:
         ep, wp, epa, wpa)
```

## Team comparison

Now using this dataset, we can proceed to compare the performance and decision 
making of the Eagles and the Patriots. First what plays did they call?

```{r play-call-counts}
super_bowl_pbp %>%
  filter(play_type %in% c("pass", "run")) %>%
  group_by(posteam, down, play_type) %>%
  count()
```

We can directly plot the proportions for each team as side-by-side barcharts

```{r play-call-bars}
super_bowl_pbp %>%
  filter(play_type %in% c("pass", "run")) %>%
  group_by(posteam, down, play_type) %>%
  count() %>%
  # Now let's remove that NA from the two point conversion:
  filter(!is.na(down)) %>%
  # Looking at counts really isn't appropriate, let's compare the proportions:
  group_by(posteam, down) %>%
  mutate(n_plays = sum(n),
         prop_plays = n / n_plays) %>%
  # Now actually create the chart!
  ggplot(aes(x = play_type, y = prop_plays, fill = posteam)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~down, ncol = 4) +
  labs(x = "Play type", y = "Proportion of plays")
```

To make this better, we can use the actual team colors based on a dataset
from the `nflscrapR` package. For the Patriots we'll use their secondary color #c60c30 
and we'll use the Eagles primary color of #004953.

```{r color-bars}
ne_color <- "#c60c30"
phi_color <- "#004953"

super_bowl_pbp %>%
  filter(play_type %in% c("pass", "run")) %>%
  group_by(posteam, down, play_type) %>%
  count() %>%
  filter(!is.na(down)) %>%
  group_by(posteam, down) %>%
  mutate(n_plays = sum(n),
         prop_plays = n / n_plays) %>%
  ggplot(aes(x = play_type, y = prop_plays, fill = posteam)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~down, ncol = 4) +
  scale_fill_manual(values = c(ne_color, phi_color)) +
  labs(x = "Play type", y = "Proportion of plays",
       fill = "Team",
       title = "Comparison of Patriots and Eagles play-calling by down") +
  theme_bw()
```

Let's take a look at the performance of these plays by yards-gained:

```{r yards-plot}
super_bowl_pbp %>%
  filter(play_type %in% c("pass", "run")) %>%
  group_by(posteam, down, play_type) %>%
  filter(!is.na(down)) %>%
  # Now use the summarise function, generate the average yards gained:
  summarize(yards_per_play = mean(yards_gained)) %>%
  ggplot(aes(x = play_type, y = yards_per_play, fill = posteam)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~down, ncol = 4) +
  scale_fill_manual(values = c(ne_color, phi_color)) +
  labs(x = "Play type", y = "Yards per play",
       fill = "Team",
       title = "Comparison of NE and PHI yards gained per play by type and down") +
  theme_bw()
```


One number summaries toss out alot of information! Let's view the entire distribution
instead. One of the best ways to do this is with a __beeswarm__ plot - which
displays the actual individual points rather than smoothed summaries. We'll display
these points on top of __violin__ plots which provide us with the general shape of
the distributions

```{r beeswarm}
# Install ggbeeswarm
# install.packages("ggbeeswarm")
library(ggbeeswarm)
super_bowl_pbp %>%
  filter(play_type %in% c("pass", "run")) %>%
  filter(!is.na(down)) %>%
  ggplot(aes(x = play_type, y = yards_gained, fill = posteam, color = posteam)) +
  geom_violin(alpha = 0.3) + 
  # Display the individual points on top of the violin plots:
  geom_beeswarm(dodge.width = 1) +
  facet_wrap(~down, ncol = 4) +
  scale_fill_manual(values = c(ne_color, phi_color)) +
  scale_color_manual(values = c(ne_color, phi_color)) +
  labs(x = "Play type", y = "Yards per play",
       fill = "Team", color = "Team",
       title = "Comparison of NE and PHI yards gained per play by type and down") +
  theme_bw() 
```

## Win probability impact

All yards are not created equal! We should really be looking at the impact in
terms of win probability added (WPA) instead to get a better understanding of
what impacted the game.

```{r wpa-dist}
super_bowl_pbp %>%
  filter(play_type %in% c("pass", "run")) %>%
  filter(!is.na(down)) %>%
  ggplot(aes(x = play_type, y = wpa, fill = posteam)) +
  geom_violin(alpha = 0.3) + 
  geom_beeswarm(aes(color = posteam), dodge.width = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray", 
             size = 1) +
  facet_wrap(~down, ncol = 4) +
  scale_fill_manual(values = c(ne_color, phi_color)) +
  scale_color_manual(values = c(ne_color, phi_color), guide = FALSE) +
  labs(x = "Play type", y = "Win probability added (WPA)",
       fill = "Team",
       title = "Comparison of NE and PHI WPA by type and down") +
  theme_bw()
```

Now we see a big difference between the Eagles and Patriots, especially on those
fourth down passing attempts... which plays are those?

```{r top-wpa-plays}
super_bowl_pbp %>%
  group_by(posteam) %>%
  filter(posteam == "PHI",
         play_type == "pass",
         !is.na(down)) %>%
  arrange(desc(wpa)) %>%
  select(desc, wpa, down, qtr) %>%
  slice(1:3)
```


The number one play was the go-ahead TD pass which gave the Eagles the lead in 
the 4th quarter, while the second highest play was the famous Philly Special!

## Win probability chart

Finally, we can wrap up this analysis with a win probability chart that shows
the overall story of the game.

```{r wp-game-chart}
# Let's put this all together to display a win probability chart:
post_pbp_2017 %>%
  # Only grab the Super Bowl
  filter(game_id == 2018020400) %>%
  filter(!is.na(home_wp),
         !is.na(away_wp),
         timeout == 0) %>%
  select(game_seconds_remaining,
         home_wp,
         away_wp) %>%
  # Rather than having separate columns for each team's win probability,
  # we can gather them into one column:
  gather(team, wpa, -game_seconds_remaining) %>%
  ggplot(aes(x = game_seconds_remaining, y = wpa, color = team)) +
  geom_line(size = 2) +
  geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") +
  scale_color_manual(labels = c("PHI", "NE"),
                     values = c(phi_color, ne_color),
                     guide = FALSE) +
  scale_x_reverse(breaks = seq(0, 3600, 300)) + 
  annotate("text", x = 3000, y = .80, label = "PHI", color = phi_color, size = 8) + 
  annotate("text", x = 3000, y = .20, label = "NE", color = ne_color, size = 8) +
  geom_vline(xintercept = 900, linetype = "dashed", color = "black") + 
  geom_vline(xintercept = 1800, linetype = "dashed", color = "black") + 
  geom_vline(xintercept = 2700, linetype = "dashed", color = "black") + 
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") + 
  labs(
    x = "Time Remaining (seconds)",
    y = "Win Probability",
    title = "Super Bowl LII Win Probability Chart",
    subtitle = "New England Patriots vs Philadelphia Eagles",
    caption = "Data from nflscrapR") + 
  theme_bw()
```

Congratulations, you now have what it takes to cover NFL games for the Athletic!

