---
title: "Problem Set 4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.align = "center", include = FALSE) # set include = TRUE to see solutions in knitted RMarkdown
library(tidyverse)
```

## Visualizing the Strike Zone

In this part, we will continue to use heatmaps (introduced briefly in [Lecture 2](lecture2.html)) to explore the strike zone in baseball. We will focus on data collected by PITCHf/x. At a high-level, PITCHf/x consists of a set of cameras installed at every ballpark which tracks the motion of each pitch. For more information about the system, check out this [article by Mike Fast](http://baseball.physics.illinois.edu/FastPFXGuide.pdf)  The data collected by PITCHf/x is then transmitted to the MLB Gameday application along with contextual information about the pitch. The dataset we'll be using contains the measurements from the PITCHf/x system recorded in 2015.

1. Download [pitchfx_2015.csv](data/pitchfx_2015.csv) to your "data" folder. Read the CSV file into a tbl called `pitches` using the `read_csv` function. 

```{r pt1 q1 solution}
# solution
pitches <- read_csv("data/pitchfx_2015.csv")
```

    The columns are:
    * `Description`: Records the outcome of the pitch (Called Strike, Swinging Strike, Foul, etc.)
    * `X` and `Z`: the horizontal and vertical coordinates of the pitch in inches. Note that the center of home plate corresponds to `X = 0`. 
      + Note that the `X` coordinate are recorded from the catcher's perspective, with negative values on the left and positive values on the right. In this coordinate system, a right-handed batter will line up to the left (i.e. negative `X` values). 
    * `COUNT`: The ball-strike count for each pitch
    * `P_HAND` and `B_HAND`: the handedness of the batter and pitcher. <p/>

2. To visualize the strike zone, we are going to want to filter out only the called strikes and balls. Moreover, it will be helpful to convert the Description to numeric values (1 for called strikes, 0 for balls). Use the pipe operator, `filter()`, `mutate()`, and `case_when()` to create a new tbl `called_pitches` containing only the called strike and balls and that includes a new column "Call" whose value is 0 for balls and 1 for called strike. 

```{r pt1 q2 solution}
#solution
called_pitches <- pitches %>%
  filter(Description == "Called Strike" | Description == "Ball") %>%
  mutate(Call = case_when(
    Description == "Called Strike" ~ 1,
    Description == "Ball" ~ 0))
```

3. To get started, we will first initialize our plot. Since we are not telling it to plot anything, it will just be blank.

```{r pt1 q3, include=TRUE}
ggplot(data = called_pitches)
```

4. To estimate the probability of a called strike given the pitch location, we will use a strategy similar to what we used to make heatmaps in [Lecture 2](lecture2.html). Essentially, we divide the plane into several small rectangular bins and compute the proportion of called strikes within each bin. To compute this, we use the `stat_summary_2d()` function, which takes three aesthetics:
    * x: variable on the horizontal axis
    * y: variable on vertical axis
    * z: variable that is passed to the summary function.
    
    By default, `stat_summary_2d()` divides the plane into rectangles based on the aesthetics x and y, and then computes the average value of z for observations in the bin. We can add this layer to our plot as follows and obtain the following plot. 
    
```{r pt1 q4, include=TRUE}
ggplot(data = called_pitches) +
  stat_summary_2d(mapping = aes(x = X, y = Z, z = Call))
```

5. You'll notice in the plot above that `stat_summary_2d()` has added a legend to our plot. However, the title of the legend is a somewhat non-informative. Moreover, the color scheme does not distinguish between different values particularly well. We can change both the title of the legend and the color scheme inside a function called `scale_fill_distiller`. Don't worry too much about what this function means for now; we will cover it in more depth in [Lecture 5](lecture5.html). 

```{r pt1 q5, include=TRUE}
ggplot(data = called_pitches) +
  stat_summary_2d(mapping = aes(x = X, y = Z, z = Call)) +
  scale_fill_distiller("P(Called Strike)", palette = "RdBu")
```

6. According to the official rule book, the strike zone is a rectangular region that spans the width of home plate and extends vertically from the batter's knee to the middle of his chest. From the plot above, we see that the region in which the strike zone probability is higher than 90% is definitely not rectangular. To better visualize the discrepancy, we can add another layer to plot which delimits an approximation of the rule book strike zone. The code below does just that. The `xmin` and `xmax` arguments give the horizontal limits of the strike zone (in this case, the coordinates of the edges of the strike zone) and the `ymin` and `ymax` arguments are the average vertical limits measured by PITCHf/x. **Note: these values were pre-computed using a much larger dataset**

```{r pt1 q6, include=TRUE}
ggplot(data = called_pitches) +
  stat_summary_2d(mapping = aes(x = X, y = Z, z = Call)) +
  scale_fill_distiller("P(Called Strike)", palette = "RdBu") +
  annotate("rect", xmin = -8.5, xmax = 8.5, ymin = 19, ymax = 41.5, alpha = 0, color = "black")

```

* We can additionally make the plot a bit more attractive visually by setting the theme to be minimal for a plain white background, removing the axis titles, and adding an overall title. 

```{r pt1 q7, include=TRUE}
ggplot(data = called_pitches) +
  stat_summary_2d(mapping = aes(x = X, y = Z, z = Call)) +
  scale_fill_distiller("P(Called Strike)", palette = "RdBu") +
  annotate("rect", xmin = -8.5, xmax = 8.5, ymin = 19, ymax = 41.5, alpha = 0, color = "black") +
  theme_minimal() + 
  theme(axis.title.x = element_blank(), 
        axis.title.y = element_blank()) + 
  labs(title = "Estimated Strike Zone")
```

## NBA Team Shooting Statistics

The file [nba_boxscore.csv](data/nba_boxscore.csv) lists detailed box score information about every NBA player in every season ranging from 1996--97 season and 2015-16 season. We will look at team shooting statistics over this 20-season span. 

1. Download the nba_boxscore data above, save it in your data folder, and add load it into a tbl called `raw_boxscore`. 

```{r pt2 q1 solution}
# solution
raw_boxscore <- read_csv("data/nba_boxscore.csv")
```

* The column "Tm" lists the team on which each player played. We can look at the relative frequencies of the teams using the `table()` function. This function takes a vector and returns the frequencies of each unique value. 

```{r pt2 q1 table, include=TRUE}
raw_boxscore %>% select(Tm) %>% table()
```

* Looking at the list of teams you may see a few that you don't recognize. For instance, there are 15 players listed as playing on "WSB". We can use `filter()` to take a closer look at these players.

```{r pt2 q1 filter, include=TRUE}
raw_boxscore %>% filter(Tm == "WSB")
```

* These fifteen players during the 1996-97 season on the Washington Bullets, which was renamed the Washington Wizards at the end of that season.There are a few other examples: VAN refers to the Vancouver Grizzlies who moved to Memphis and CHH refers to the original Charlotte Hornets franchise, which ultimately relocated to New Orleans.

* One of the teams listed is "TOT". This does not refer any specific team. Instead these rows record the *total* statistics recorded by a player if he played for multiple teams in a single season. For the purposes of understanding how team shooting statistics changed over time, we will not want to include these rows in our analysis.

2. Use `filter()`, `group_by()`, `reframe()`, and `mutate()` to create a new tbl called `team_boxscore` that does the following:

    * removes the rows corresponding to player totals (i.e. `Tm == "TOT"`)
    * groups the tbl according to Season and Tm. **Note: it is important to group on season first and team second **
    * Computes the total number of made and attempted field goals, three pointers, and free throws, along with points scored by each team in each season. 
    * Adds a column for team field goal percentage (FGP), three point percentage (TPP), and free throw percentages (FTP). 
    * Ungroup the resulting tbl <p/>

```{r pt2 q2 solution}
# solution
team_boxscore <-
  raw_boxscore %>% 
  filter(Tm != 'TOT') %>%
  group_by(Season, Tm) %>% 
  reframe(total_FGM = sum(FGM), 
            total_FGA = sum(FGA),
            total_TPM = sum(TPM), 
            total_TPA = sum(TPA),
            total_FTM = sum(FTM), 
            total_FTA = sum(FTA),
            total_PTS = sum(PTS)) %>%
  mutate(FGP = total_FGM/total_FGA,
         TPP = total_TPM/total_TPA,
         FTP = total_FTM/total_FTA) %>%
  ungroup()
```

3. Use `filter()` to create a new tbl called `reduced_boxscore` that pulls out the rows of `team_boxscore` corresponding to the following teams: BOS, CLE, DAL, DET, GSW, LAL, MIA, and SAS. Then create a line plot of these teams' three point percentage in each season. Be sure to color the points according to the team (*Hint*: to map the `Tm` variable to the points as colors, use the `color =` argument within the `aes` function. We will learn more about this in [Lecture 5](lecture5.html)). What patterns do you notice?

```{r pt2 q3 solution}
# solution
reduced_boxscore <-
  team_boxscore %>%
  filter(Tm %in% c('BOS', 'CLE', 'DAL', 'DET', 'GSW', 'LAL', 'MIA', 'SAS'))

ggplot(data = reduced_boxscore) +
  geom_line(mapping = aes(x = Season, y = TPP, color = Tm), size = 1) 
```

4. Use `pivot_long()` on `reduced_boxscore` to create separate rows for each team's FGP, TPP, and FTP. Create a tibble filtered on one team only and visualize how their FGP, TPP, and FTP have all evolved over time on one plot. *Challenge: use `facet_wrap()` with teams on the entire `reduced_boxscore` table to visualize the percentages for all 8 teams at once. 
```{r pt2 q4 solution}
boxscore_long <- reduced_boxscore %>%
  pivot_longer(cols = c(FGP, TPP, FTP), 
               names_to = "Statistic", 
               values_to = "Percentage")

ggplot(boxscore_long, aes(x = Season, y = Percentage, color = Statistic)) +
  geom_line() +
  labs(title = "Team Shooting Percentages Over Seasons",
       x = "Season",
       y = "Percentage") +
  theme_minimal() +
  facet_wrap(~Tm) +
  scale_color_manual(values = c("FGP" = "palegreen3", "TPP" = "lightblue", "FTP" = "lightcoral"))
```


## Work with new data

Once you finish reviewing the material from earlier this week, we'd like you to use some of the tools we introduced in [Lecture 4](lecture4.html) to read data into R.

Choose from some of the following datasets to take a look at:

* [Ironman data.csv](data/Ironman Data.csv)
* [bad-drivers.csv](data/bad-drivers.csv)
* [batting_tto.csv](data/batting_tto.csv)

1. Load in the one of the datasets and inspect their features with the `head()` function.
```{r read in data}
batting_tto <- read_csv("data/batting_tto.csv")
```

2. Try to make some visualizations with ggplot that explore the data. For example, plotting batting time through the order versus their event wOBA for the pitch. If you are working with the TTO data, try exploring something other than this for your problem set. 

```{r visualizations}
ggplot(batting_tto, aes(x = ORDER_CT, y = EVENT_WOBA_19)) +
  geom_point(alpha = 0.5) +
  labs(title = "Batter TTO vs Event wOBA",
       x = "Batter TTO",
       y = "Event wOBA") +
  theme_minimal()
```
3. Next, try mutating your data with `reframe()`, utilizing `pivot()`, or computing correlation with `cor()` to generate another visualization. For example, plotting the mean event wOBA with `geom_col()`.

```{r grouping}
batting_tto_grouped <- batting_tto %>% 
  group_by(ORDER_CT) %>%
  reframe(mean_wOBA = mean(EVENT_WOBA_19, na.rm = TRUE))
```


```{r geom col, include=TRUE}
ggplot(batting_tto_grouped, aes(x = ORDER_CT, y = mean_wOBA)) +
  geom_col(fill = "lightblue") +
  labs(title = "Mean Event wOBA by Batter Sequence",
       x = "Batter TTO",
       y = "Mean Event wOBA") +
  theme_minimal()
```

4. Try computing some averages with `reframe()` and test their predictive power on the dataset!
