---
title: "Problem Set 7"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.align = "center", include = FALSE) # set include = TRUE to see solutions in knitted RMarkdown
library(tidyverse)
library(ggplot2)
```


# Testing Regression Models

We will now continue modeling data with regression, but we can now test and compare different models more efficiently. We will use this * [mlb_payrolls_train.csv](data/mlb_payrolls_train.csv) to analyze general manager performance based on team payrolls and winning percentages.

1. First, load in the data to R. Let's see the range of seasons that are included in this data 
```{r, include = TRUE}
payroll <- read_csv("data/mlb_payrolls_train.csv")

unique(payroll$yearID)
```
1. We can see that the data ranges from 1998 to 2023. Since the 2020 season was shortened due to Covid, filter it out. Then use `mutate()` to create a column names `WP` based on wins (W) and games (G). 

```{r column names}
head(payroll)
```


```{r filter and mutate}
payroll <- payroll %>% 
  filter(yearID != 2020) %>% 
  mutate(WP = W/G)
```


2. Then, explore the relationship between payrolls and winning percentage by generating a scatterplot. Use `ifelse()` with the `%in%` operator to mark the Oakland A's and New York Yankees' points on the plot. Be sure to mark them with different colors by using `scale_color_manual()` and to change the alpha values if needed so that overlapping point's don't completely obscure each other.
```{r plot}
ggplot(payroll, aes(x = payroll_norm, y = WP, 
                    color = ifelse(name %in% c("New York Yankees", "Oakland Athletics"), name, "Other"))) +
  geom_point(size = 3, alpha = 0.5) +
  labs(
    title = "MLB Payrolls vs Winning Percentage",
    x = "Normalized Payroll",
    y = "Winning Percentage",
    color = "Team"
  ) +
  theme_minimal() +
  scale_color_manual(values = c(
    "New York Yankees" = "navy",
    "Oakland Athletics" = "green",
    "Other" = "black"
  ))

```

3. Now, fit a regression line to the data as a function of `Norm_Payroll`, then add it to the plot. 
```{r}
payroll_model <- lm(WP ~ payroll_norm, data = payroll)

payroll <- payroll %>%
  mutate(predicted_wp = predict(payroll_model))

# Plot with predicted line
ggplot(payroll, aes(x = payroll_norm, y = WP)) +
  geom_point() +
  geom_line(aes(y = predicted_wp), color = "lightcoral", size = 1) +
  labs(
    title = "MLB Payrolls vs Winning Percentage (with Regression Line)",
    x = "Normalized Payroll",
    y = "Winning Percentage"
  ) +
  theme_minimal()
```

4. Next, try fitting a model based on `log_payroll_norm` and adding that to the plot. 
```{r}
log_model = lm(WP ~ (log_payroll_norm), data = payroll)
payroll <- payroll %>%
  mutate(log_predicted_wp = predict(log_model))
ggplot(payroll, aes(x = payroll_norm, y = WP)) +
  geom_point() +
  geom_line(aes(y = log_predicted_wp), color = "lightcoral", size = 1) +
  labs(
    title = "MLB Payrolls vs Winning Percentage (with Regression Line)",
    x = "Normalized Payroll",
    y = "Winning Percentage"
  ) +
  theme_minimal()
```
You should see that the log model will be non-linear. Just by looking at the plots, which seems to fit the data better? 

5. One way to potentially look at model performance is with the residuals. Using `add_residual` from `modelr`, add a column of residuals from each model to your payroll table. Then, generate a scatter plot of the residuals for each model. 
```{r}
library(modelr)
payroll <- payroll %>%
  add_residuals(payroll_model, var = "residuals") %>%
  add_residuals(log_model, var = "log_residuals")

ggplot(payroll, aes(x = payroll_norm)) +
  geom_point(aes(y = residuals), color = "lightblue") +
  geom_point(aes(y = log_residuals), color = "lightcoral") +
  labs(
    title = "Residuals of Payroll Models",
    x = "Normalized Payroll",
    y = "Residuals"
  ) +
  theme_minimal() 
```

An indication of a poor model fit is a non-uniform residual plot. This would indicate that perhaps a linear model is being used when the underlying trend is instead logarithmic, or polynomial. Do you observe this? 

6. Finally, we can directly compare our models with out-of-sample performance. Load in  *[mlb_payrolls_test.csv](data/mlb_payrolls_test.csv) and call it `payrolls_test`. Do not refit your model but instead use `predict` to generate predictions for the test data set called `test_pred` and `test_log_pred`, similar to its usage in Lecture 6. Then, use the line below to compute the RMSE for each model. 
```{r}
payrolls_test <- read_csv("data/mlb_payrolls_test.csv")
payrolls_test <- payrolls_test %>%
  mutate(
    test_pred = predict(payroll_model, newdata = payrolls_test),
    test_log_pred = predict(log_model, newdata = payrolls_test)
  )
```

```{r RMSE, include = TRUE}
rmse <- sqrt(mean((payrolls_test$WP - payrolls_test$test_pred)^2))
rmse_log <- sqrt(mean((payrolls_test$WP - payrolls_test$test_log_pred)^2))
```

```{r values}
rmse
rmse_log
```

Based on the RMSE values, which model is better? 
