---
title: "Problem Set 8"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.align = "center", include = FALSE) # set include = TRUE to see solutions in knitted RMarkdown
library(tidyverse)
library(ggplot2)
```


# Logistic Regression

We will now continue modeling data with logistic regression using * [nfl_wp_train.csv](data/nfl_wp_train.csv) to analyze win percentages as a function of multiple variables. The dataset includes `game_id`, `season`, `label_win`, which is a binary (0 or 1) indicator on whether the possessing team 


```{r}
plot(cars)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

