---
title: "Problem Set 8"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.align = "center", include = FALSE) # set include = TRUE to see solutions in knitted RMarkdown
library(tidyverse)
library(ggplot2)
```


# Logistic Regression

We will now continue modeling data with logistic regression using * [nfl_wp_train.csv](data/nfl_wp_train.csv) to analyze win percentages as a function of multiple variables. The dataset includes `game_id`, `season`, `label_win`, which is a binary (0 or 1) indicator on whether the possessing team won the game, `score_differential`, which is the current score differential between the team in possession and the team on defense, `posteam_spread`, which is the +- spread for the possessing team prior to the game start, `yardline_100`, `down`, and `ydstogo`. 

1. First, load in the data. Next, do some exploratory data analysis (EDA) to understand the data. For example, you can plot the distribution of `score_differential` and `label_win`, and create a scatter plot of `score_differential` vs. `label_win`. Since the target variable, `label_win`, is binary, a good way to visualize the data would be to group data based on bins in your chosen variable and then calculate the win percentage based on the number of `label_win` = 1 values divided by the total number of observations in each bin.


2. Then fit a logistic model using `glm` to predict `label_win` based on `score_differential`, then add the model's predictions to the dataset.
```{r model and predictions.}
nfl_wp = read.csv('data/nfl_wp_train.csv')
spread_model = glm(label_win ~ score_differential, data = nfl_wp)
spread_model
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

