---
title: 'Lecture 2: The grammar of graphics and filtering data'
output: html_document
---

```{r lecture2-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "", prompt = TRUE, collapse = TRUE, tidy=FALSE, fig.align = "center")
library(tidyverse)
```

We're going to pick up right where we left off yesterday in [Problem Set 1](tc_ps1.html), where you wrote a script to create a `tbl` containg several shooting statistics for NBA players between the 1996-97 season and the 2015-16 season.
Your script could look something like what is the code block below, and for reasons that will become clear shortly, we are going to call our `tbl` `raw_shooting` (remember: writing code in script files is good practice!)
```{r ps2-create-tbl, tidy=FALSE}
library(tidyverse)
# Read in data
raw_shooting <- read_csv(file = "data/nba_shooting.csv")
# Create new columns
raw_shooting <-
  mutate(raw_shooting,
         FGP = FGM / FGA,
         TPP = TPM / TPA,
         FTP = FTM / FTA,
         eFGP = (FGM + 0.5 * TPM) / (FGA),
         PTS = FTM + 2 * FGM + TPM,
         TSP = PTS/(2 * (FGA + 0.44 * FTA)))
# Sort by the TSP in descending order
raw_shooting <- arrange(raw_shooting, desc(TSP))
```

As we learned at the end of [Lecture 1](tc_lecture1.html), rather than repeatedly 
referring to our `tbl`, we can use the pipe `%>%` to chain together our operations
for a much _cleaner_ chunk of code:

```{r ps2-create-tbl-pipe, tidy=FALSE}
# First read in the data, assigning pipeline to raw_shooting
raw_shooting <- read_csv(file = "data/nba_shooting.csv") %>%
  # Next create the new columns
  mutate(FGP = FGM / FGA,
         TPP = TPM / TPA,
         FTP = FTM / FTA,
         eFGP = (FGM + 0.5 * TPM) / (FGA),
         PTS = FTM + 2 * FGM + TPM,
         TSP = PTS / (2 * (FGA + 0.44 * FTA))) %>%
  # And finally sort by TSP
  arrange(desc(TSP)) 
```


## Introduction to ggplot2

We will use visualizations to answer some questions about the data. Specifically, we will study the distribution of the individual columns as well as try to understand the relationship between pairs of variables. For instance, are players attempting and making more three point shots now than they did 10 years ago?

Throughout the week, we will be using the popular [`ggplot2`](https://ggplot2.tidyverse.org) package (again created by Hadley Wickham and a member of the `tidyverse`) for all of our data visualizations. The `gg` stands for the **grammar of graphics**, an intuitive framework for data visualization. Given a dataset, such as `raw_shooting`, we want to map the columns to certain __aesthetics__ of a visualization, such as the x-axis, y-axis, size, color, etc. Then a geometric
object is used to represent the aesthetics visually such as a barchart or scatterplot. This framework separates the process of visualization into different components: data, aesthetic mappings, and geometric objects. These components are then added together (or __layered__) to produce the final graph. The `ggplot2` package is the most popular implementation of the __grammar of graphics__ and is relatively easy to use. 

As you saw in Prof. Wyner's lectures, histograms are a powerful way to describe a single dataset. Let's start with making a histogram of field-goal percentage (FGP). To do so with `ggplot2`, we  start by telling `R` which `tbl` we want to use as the data for the plot. This is done using the `ggplot()` function:

```{r shooting_hist}
ggplot(data = raw_shooting)
```

As you can see, nothing is displayed! That's because we've just initialized the dataset
to be used for creating the plot. Next, we need to map variables/columns from the data to __aesthetics__/properties of the plot. Examples include:

* `x`: the variable that will be on the x-axis
* `y`: the variable that will be on the y-axis
* `color`: the variable that categorizes data by color
* `shape`: the variable that categorizes data by shape

For the histogram, we map FGP to the `x` aesthetic with the `aes()` functions, displaying values of FGP along
the x-axis:

```{r shooting_hist_x}
ggplot(data = raw_shooting, aes(x = FGP))
```

Now we can see an axis for FGP, but still no histogram! That's because we need to __add__ the geometric layer of the histogram to the plot. The various geometric objects available in `ggplot2` are referred to as `geom`s, and these ultimately determine the type of play that will be created. Examples include:

* `geom_point()`: creates a scatterplot
* `geom_histogram()`: creates a histogram
* `geom_line()`: creates a line
* `geom_boxplot()`: creates a boxplot

To display the FGP histogram, we will simply add the `geom_histogram()` layer to the
plot using the `+` operator:

```{r shooting_hist_x_geom, tidy = FALSE}
ggplot(data = raw_shooting, aes(x = FGP)) +
  geom_histogram()
```

And of course, we can equivalently generate the same figure by __piping__ our dataset
into the ``ggplot`` function. The remaining examples will use the pipe operator `%>%` 
for the remainder of this week, emphasizing how you can manipulate the `tbl` with various
other functions prior to displaying the data:

```{r pipe_shooting_hist_x_geom, tidy = FALSE}
raw_shooting %>%
  ggplot(aes(x = FGP)) +
  geom_histogram()
```

Each `geom` has it various attributes that can be modified whether by mapping variables
with `aes` or _global_ settings that affects the final displayed plot. A common problem
faced with histograms is determining the width of the bins or number of bins - how much
smoothing do we want of the data? You might have noticed that by default `geom_histogram` 
is using 30 bins and prints out a message for us about this decision. We can easily
modify the number of bins in the `geom_histogram` function using the `bins` argument:

```{r shooting_hist_nbins, tidy = FALSE}
raw_shooting %>%
  ggplot(aes(x = FGP)) +
  geom_histogram(bins = 50)
```

Or we directly specify the width of the bins reflecting the range of FGP included
in each bin using `binwidth`:

```{r shooting_hist_binwidth, tidy = FALSE}
raw_shooting %>%
  ggplot(aes(x = FGP)) +
  geom_histogram(binwidth = 0.05)
```

**Now you should spend time making histograms of FTP and TPP, and other variables created in code above. Discuss the differences with others.**

## Filtering

By this point, it should be clear that there are a number of rather curious features in our dataset.
For instance, there seem to be several players who have never made a field goal but have made every one of their free throw attempts. As it turns out, we have several players who have attempted fewer than 5 field goals in a single season.
We'd like to remove all of the players who have not attempted a considerable number of field goal attempts and three point attempts in order to understand how the rate and efficiency of three point shots has changed over time.

The `filter()` function is used to pull out subsets of observations that satisfy some logical condition like `FGA > 100` or `FGA > 100` and `FTA > 50`.
To make such comparisons in R, we have the following operators available at our disposal:

* `==` for "equal to"
* `!=` for "not equal to"
* `<` and `<=` for "less than" and "less than or equal to"
* `>` and `>=` for "greater than" and "greater than or equal to"
* `&`, `|`, `!` for "AND" and "OR" and "NOT"

The code below filter out all of the players with at least 100 field goals in a single season
```{r filter-fga}
raw_shooting %>%
  filter(FGA > 100)
```
When we run this code, you'll notice that R prints out a `tbl` with 6,295 rows.

We can also filter on more complicated conditions constructed using the AND, OR, and NOT operators: `&`, `|`, and `!`.
For instance, to filter observations with at least 100 field goal attempts OR 50 three point attempts, we would do
```{r filter-or}
raw_shooting %>%
  filter(FGA >= 100 | TPA >= 50)
```

We may combine these constraints by enclosing them in parantheses.
```{r filter-complex}
raw_shooting %>%
  filter((FGA >= 100 & TPA >= 50) | (FGP >= 0.45 & FGP <= 0.5))
```

What if we wanted to pull out the observations corresponding to the 2015-16 and 2014-15 season?
We could do something like `filter(raw_shooting, (SEASON == 2016) | (SEASON == 2015))`, which would be perfectly fine.
However, what if we wanted data from 1998-99, 2011-12, and 2015-16? 
Typing a lot of expressions like `SEASON == ...` would be rather tedious.
The `%in%` operator lets us avoid this tedium: 
```{r filter-in}
raw_shooting %>%
  filter(SEASON %in% c(1999, 2012, 2016))
```

We could also filter out data from the two lockout-shortened seasons, 1998-99 and 2011-12 using a combination of the NOT `!` operator and `%in%`.
```{r filter-not-in}
raw_shooting %>%
  filter(!SEASON %in% c(1999, 2012))
```

For the remainder of this module, we will focus on the non-lockout seasons (ie `!SEASON %in% c(1999, 2012, 2016)`) but we still need to determine a cutoff for FGA and TPA. Let's start by making histograms of the two variables to see their individual distributions, and rather than creating a new temporary `tbl` we'll take advantage of the `%>%`. First for FGA,

```{r fga-hist, tidy = FALSE}
raw_shooting %>%
  filter(!SEASON %in% c(1999, 2012)) %>%
  ggplot(aes(x = FGA)) +
  geom_histogram(binwidth = 50)
```

And now for TPA,

```{r tpa-hist, tidy = FALSE}
raw_shooting %>%
  filter(!SEASON %in% c(1999, 2012)) %>%
  ggplot(aes(x = TPA)) +
  geom_histogram(binwidth = 25)
```

It might not make sense however to make a cutoff for FGA and TPA without considering
their relationship. 

## Bivariate plots for continuous data

We can proceed to view the __joint__ distribution of FGA and TPA, the relationship
between the two variables, by displaying a scatterplot of the data. Obviously, as
TPA increases the FGA will increase (since FGA is the sum of TPA and number of two-point
attempts). We create a scatterplot by mapping multiple variables to `x` and `y`,
and use `geom_point` to display the desired geometric object of points instead of
`geom_histogram` from before.

```{r scatter-fga-tpa, tidy = FALSE}
raw_shooting %>%
  filter(!SEASON %in% c(1999, 2012)) %>%
  ggplot(aes(x = TPA, y = FGA)) +
  geom_point()
```

Immediately, we notice a few things about the figure. First, we see the clear relationship between TPA and FGA - where where TPA provides the lower threshold on FGA (remember why is this expected). We also see what appears to be a group of points with a smaller number of three-point attempts displaying a range in values for FGA, while a main block of points shows an clear increasing relationship. A major drawback of this scatterplot is its inability to show the relative density of points. For instance, all we see is a solid black mass for the lower range of values for both TPA and FGA - making it hard to determine where an appropriate cutoff should be made.

One way to address this is to use *alpha-blending* to change the transparency of each point. 
When there are many points plotted in the same region, that region will appear darker. Just like
`binwidth` or `bins` for `geom_histogram`, `geom_point` has settings we can change such as the `alpha`
setting in this case which ranges from 0 (completely transparent) to 1 (solid and opaque, the default).

```{r scatterplot-alpha, tidy = FALSE}
raw_shooting %>%
  filter(!SEASON %in% c(1999, 2012)) %>%
  ggplot(aes(x = TPA, y = FGA)) +
  geom_point(alpha = 0.1)
```

Now we have a much better idea of where the majority of points are, with a clear group of players displaying a smaller number of TPA. Another type of plot commonly used in this situation is a __heatmap__ which you can thnk of as a two-dimensional histogram. To form a heatmap, you start by dividing the coordinate plane into many evenly-sized two-dimensional bins and then count the number of points within each bin. You then color the bin according to the count. 
While you can conceptually make the bins any shape you want, there are two popular conventions: rectangular binning and hexagonal binning. For this plot, we will focus on rectangular binning, using the `geom_bin2d()` function.
```{r heatmap, tidy = FALSE}
raw_shooting %>%
  filter(!SEASON %in% c(1999, 2012)) %>%
  ggplot(aes(x = TPA, y = FGA)) +
  geom_bin2d()
```

Now we have a color scale that has appeared to tell us the number of points in each bin. Just like histograms, we can increase the number of bins to get a much more high-resolution view of our data.
```{r, tidy = FALSE}
raw_shooting %>%
  filter(!SEASON %in% c(1999, 2012)) %>%
  ggplot(aes(x = TPA, y = FGA)) +
  geom_bin2d(bins = 100)
```

In this example, using `geom_point` with lower values for `alpha` provides a clear
interpretation for where to make the cutoff. Given the histograms we previously
made for each variable as well, let's use a cutoff of `TPA > 50` and `FGA > 150`. 
We can create the scatterplot as before but, to demonstrate a convenient feature
of `ggplot2`, we'll now assign the plot to a variable named `fga_tpa_plot`,

```{r init-fga-tpa-plot, tidy = FALSE}
fga_tpa_plot <- raw_shooting %>%
  filter(!SEASON %in% c(1999, 2012)) %>%
  ggplot(aes(x = TPA, y = FGA)) +
  geom_point(alpha = 0.1)
```

Notice how this code runs without displaying the plot. To display it we simply
run `fga_tpa_plot` in the console and the plot appear,

```{r}
fga_tpa_plot
```

Now we're going to annotate this plot with the cutoffs we've determined. To do so,
we're going to add a vertical line to provide the cutoff for the minimum value of TPA,
and a horizontal line for the minimum value of FGA. Both of these can be accomplished
using the `geom`s `geom_vline` and `geom_hline` where we specify for each the values
for the intercepts to draw the lines at. Since we've stored the plot in `fga_tpa_plot`,
we can add these layers with the `+` operator to `fga_tpa_plot` directly:

```{r add-lines, tidy = FALSE}
fga_tpa_plot +
  # Add TPA cutoff
  geom_vline(xintercept = 50) +
  # Add FGA cutoff
  geom_hline(yintercept = 150)
```

This provides us with a good indication of what we're cutting off, but we should
distinguish these lines separately from the points more. Since these are both
`geom`s with their own set of attributes, we'll change the color and make the line
type to be dashed,

```{r add-lines-color, tidy = FALSE}
fga_tpa_plot +
  # Add TPA cutoff
  geom_vline(xintercept = 50, color = "red", linetype = "dashed") +
  # Add FGA cutoff
  geom_hline(yintercept = 150, color = "red", linetype = "dashed")
```

Although `ggplot2` automatically provides axis labels based on the variables we've
mapped to the aesthetics in `aes`, we really need text describing what is shown
in the plot. The easiest way to do this, is by adding a label layer with the `labs()` function.
Using `labs`, you can provide it the same arguments as those inside `aes` as well as
other parts of the plot to label such as the title, subtitle, and caption. We add
appropriate labels to the plot above, with better descriptions for the axes, as
well as an informative subtitle regarding the red-dashed lines:

```{r add-labels, tidy = FALSE}
fga_tpa_plot +
  # Add TPA cutoff
  geom_vline(xintercept = 50, color = "red", linetype = "dashed") +
  # Add FGA cutoff
  geom_hline(yintercept = 150, color = "red", linetype = "dashed") +
  # Add appropriate labels:
  labs(x = "Number of three point attempts (TPA)",
       y = "Number of field goal attempts (FGA)",
       title = "Scatterplot of FGA and TPA (excluding lockout seasons 1999 and 2012)",
       subtitle = "Red-dashed lines indicate cutoffs for TPA > 50 and FGA > 150",
       caption = "Created by INSERT YOUR NAME HERE")
```

As a reminder, the code chunk above is equivalent to running the entire pipeline
of code without creating any temporary objects, although not recommended...

```{r full-pipeline, tidy = FALSE}
read_csv(file = "data/nba_shooting.csv") %>%
  filter(!SEASON %in% c(1999, 2012)) %>%
  ggplot(aes(x = TPA, y = FGA)) +
  geom_point(alpha = 0.1) +
  geom_vline(xintercept = 50, color = "red", linetype = "dashed") +
  geom_hline(yintercept = 150, color = "red", linetype = "dashed") +
  labs(x = "Number of three point attempts (TPA)",
       y = "Number of field goal attempts (FGA)",
       title = "Scatterplot of FGA and TPA (excluding lockout seasons 1999 and 2012)",
       subtitle = "Red-dashed lines indicate cutoffs for TPA > 50 and FGA > 150",
       caption = "Created by INSERT YOUR NAME HERE")
```

## Saving our work

For next [lecture](tc_lecture3.html) we will focus on the dataset of players who attempted at least 150 field goals and 50 three pointers in the non-lockout seasons. We'll make this _clean_ dataset using the
`filter()` function separating each condition with commas, and refer to
the resulting dataset as `nba_shooting`. Finally, we'll add a `mutate()` line to
create the `three_point_fg_rate = TPA / FGA` variable from [yesterday](tc_lecture1.html).

```{r create-nba_shooting, tidy = FALSE}
nba_shooting <- raw_shooting %>%
  # Filter on the conditions
  filter(!SEASON %in% c(1999, 2012),
         TPA > 50,
         FGA > 150) %>% 
  # Create the three point attempt rate variable
  mutate(three_point_fg_rate = TPA / FGA)
```

While we can always re-run the commands used to produce this `tbl` from our script, when data analyses become more complicated, it is helpful to save these objects.
`R` has its own special file format for efficiently saving data on your computer.

We will use the `save()` command.
```{r save}
save(nba_shooting, file = "data/clean_nba_shooting.RData")
```

Then when we want to *load* the data back into R, we can use the `load()` function
```{r load}
load("data/clean_nba_shooting.RData")
```

Next, proceed to practice this lecture's lessons in [Problem Set 2](tc_ps2.html).
