---
title: "Problem Set 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Review Lecture 2 Notes

Please spend a few minutes reading through the notes from [Lecture 2](tc_lecture2.html). 
Like in [Problem Set 1](tc_ps1.html), you should go through each code block with someone in your group and see if you can both explain to each other what all of the code does.

## Payroll and Winning Percentage in the MLB

In lecture, Prof. Wyner discussed the relationship between a team's payroll and its winning percentage.
We will create plots from his analysis in the following problems using the dataset "mlb_relative_payrolls.csv", which you can find in the "data/" folder of your working directory.
You should save all of the code for this analysis in an R script called "ps2_mlb_payroll.R".

If you don't have the dataset in your data folder yet, download it from the GitHub page [here](https://github.com/skdeshpande91/wharton_moneyball/raw/master/website_files/data/mlb_relative_payrolls.csv).

1. Read the data in from the file and save it as a tbl called "relative_payroll"
```{r read-payrolls}
relative_payroll <- read_csv(file = "data/mlb_relative_payrolls.csv")
```

2. Make a histogram of team winning percentages. Play around with different binwidths. 
```{r, eval = FALSE, echo = FALSE}
ggplot(data = relative_payroll) +
  geom_histogram(mapping = aes(x = Winning_Percentage, y = ..density..), binwidth = 0.01)
```

3. Make a histogram of the relative payrolls.
```{r, eval = FALSE, echo = FALSE}
ggplot(data = relative_payroll) +
  geom_histogram(mapping = aes(x = Relative_Payroll, y = ..density..), binwidth = 0.1)
```

4. Make a scatterplot with relative payroll on the horizontal axis and winning percentage on the vertical axis. 

```{r, eval = FALSE, echo = FALSE}
ggplot(data = relative_payroll) +
  geom_point(mapping = aes(x = Relative_Payroll, y = Winning_Percentage))
```

5. Without executing the code below, discuss with your group and see if you can figure out what it is doing.
```{r, eval = FALSE}
relative_payroll %>%
  ggplot(aes(x = Year, y = Team_Payroll)) +
  geom_point()
```

6. Execute the code above. What can you say about how team payrolls have evolved over time? Make a similar plot that visualizes how *relative payrolls* have evolved over time.

```{r, eval = FALSE, echo = FALSE}
ggplot(data = relative_payroll) + 
  geom_point(mapping = aes(x = Year, y = Relative_Payroll))
```

7. Add an appropriate title and relabel the y-axis using the `labs()` function to the plot above.
