---
title: "Problem Set 3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.align = "center", include = FALSE) # set include = TRUE to see solutions in knitted RMarkdown
library(tidyverse)
```

## Review Lecture 3 Notes

Please spend a few minutes reading through the notes from [Lecture 3](tc_lecture3.html).  Like before, you should go through each code block with someone in your group and see if you can both explain to each other what all of the code does.

## NBA Three-Point Efficiency

In [Lecture 3](tc_lecture3.html) we saw that three-point shot rates have been increasing over time by creating visualizations of the distributions for each season. 

Now, create the same visuals but use TPP as the continuous variable of interest. As players are attempting more three-pointers than ever before, are they also making them more?

Discuss with other students in your group the meaning of these plots.

```{r q1 solution}
# solution

# load data
load("data/clean_nba_shooting.RData")

# scatterplot of 3PT proportion by year
nba_shooting %>%
  ggplot(aes(x = SEASON, y = TPP)) +
  geom_point(alpha = 0.25) +
  labs(x = "Season",
       y = "3PT proportion = TPM / TPA",
       title = "Distribution of 3PT proportion each year")

# add season_factor variable
nba_shooting <- nba_shooting %>%
  mutate(season_factor = as.factor(SEASON))

# many histograms of 3PT proportion by year
nba_shooting %>%
  ggplot(aes(x = TPP)) +
  geom_histogram(binwidth = 0.05) +
  labs(x = "3PT propotion = TPM / TPA",
       y = "Count",
       title = "Distribution of 3PT proportion by season") +
  facet_wrap(~season_factor, nrow = 3)

# stacked histogram of 3PT proportion, colored by year
nba_shooting %>%
  ggplot(aes(x = TPP, fill = season_factor)) +
  geom_histogram(binwidth = 0.05) +
  labs(x = "3PT proportion = TPM / TPA",
       y = "Count",
       fill = "Season",
       title = "Distribution of 3PT proportion by season") 

# overlaid histograms of 3PT proportion, colored by year
nba_shooting %>%
  ggplot(aes(x = TPP, fill = season_factor)) +
  geom_histogram(binwidth = 0.05, position = "identity", alpha = 0.1) +
  labs(x = "3PT proportion = TPM / TPA",
       y = "Count",
       fill = "Season",
       title = "Distribution of 3PT proportion by season") 

# boxplots of 3PT proportion by year
nba_shooting %>%
  ggplot(aes(x = season_factor, y = TPP)) +
  geom_boxplot() +
  labs(x = "Season",
       y = "3PT proportion = TPM / TPA",
       title = "Comparison of 3PT proportion for each season with boxplots") 

# violin plots of 3PT proportion by year
nba_shooting %>%
  ggplot(aes(x = season_factor, y = TPP)) +
  geom_violin() +
  # label!
  labs(x = "Season",
       y = "3PT proportion = TPM / TPA",
       title = "Comparison of 3PT proportion for each season with violin plots") 

# box/violin plots of 3PT proportion by year
nba_shooting %>%
  ggplot(aes(x = season_factor, y = TPP)) +
  geom_violin() + 
  geom_boxplot(width = 0.2) +
  # label!
  labs(x = "Season",
       y = "3PT proportion = TPM / TPA",
       title = "Comparison of 3PT proportion for each season with violin and boxplots") 

# ridgeplots of 3PT proportion by year
library(ggridges)
nba_shooting %>%
  ggplot(aes(x = TPP, y = season_factor)) +
  geom_density_ridges() +
  # label!
  labs(y = "Season",
       x = "3PT proportion = TPM / TPA",
       title = "Comparison of 3PT proportion for each season with ridges") 

# beeswarm plots of 3PT proportion by year
library(ggbeeswarm)
nba_shooting %>%
  ggplot(aes(x = season_factor, y = TPP)) +
  geom_quasirandom(size = 0.5) +
  # label!
  labs(x = "Season",
       y = "3PT proportion = TPM / TPA",
       title = "Comparison of 3PT proportion for each season with beeswarms") 

# group the data by season_factor
nba_shooting <- nba_shooting %>% 
  group_by(season_factor)

# compute average 3PT proportion for each season
nba_shooting %>%
  reframe(ave_TPP = mean(TPP))

# new dataset with mean and SD of 3PT proportion each year
nba_shooting_summary <- nba_shooting %>%
  reframe(ave_TPP = mean(TPP),
            sd_TPP = sd(TPP))

# scatterplot of mean +/- SD for 3PT proportion each year 
nba_shooting_summary %>%
  ggplot(aes(x = season_factor)) +
  geom_point(aes(y = ave_TPP)) +
  geom_errorbar(aes(ymin = ave_TPP - sd_TPP,
                    ymax = ave_TPP + sd_TPP)) +
  labs(x = "Season",
       y = "3PT proportion = TPM / TPA",
       title = "Average 3PT proportion with +/- 1 standard deviation")

# violin plot plus mean +/- SD for 3PT proportion each year
ggplot() +
  geom_violin(data = nba_shooting,
              aes(x = season_factor, y = TPP)) +
  geom_point(data = nba_shooting_summary,
             aes(x = season_factor, y = ave_TPP),
             color = "red") +
  geom_errorbar(data = nba_shooting_summary,
                aes(x = season_factor, 
                    ymin = ave_TPP - sd_TPP,
                    ymax = ave_TPP + sd_TPP),
                color = "red") +
  labs(x = "Season",
       y = "3PT proportion = TPA / FGA",
       title = "Distribution of 3PT proportion for each season",
       subtitle = "Red points indicate average for each year with +/- 1 standard deviation")

# beeswarm plot plus mean +/- SD for 3PT proportion each year
ggplot() +
  geom_quasirandom(data = nba_shooting,
                   aes(x = season_factor, y = TPP),
                   size = 0.5) +
  geom_point(data = nba_shooting_summary,
             aes(x = season_factor, y = ave_TPP),
             color = "red") +
  geom_errorbar(data = nba_shooting_summary,
                aes(x = season_factor, 
                    ymin = ave_TPP - sd_TPP,
                    ymax = ave_TPP + sd_TPP),
                color = "red") +
  labs(x = "Season",
       y = "3PT proportion = TPA / FGA",
       title = "Distribution of 3PT proportion for each season",
       subtitle = "Red points indicate average for each year with +/- 1 standard deviation")
```


## Other Data Sources

As you start to brainstorm ideas for your projects, consider some of the R packages that provide interesting datasets from various sports and additional functions for working with that data. For more information, see the [Data Sources Lecture](tc_lecture_data_sources.html).